Guarded Recursion in Agda via Sized Types

### 2019-01-17 code-oneclock-paper/latex/Introduction.tex
WIP intro

Dependent type theory is a very expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda \cite{norell2008} and Lean.  It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Elements of these types, which we call processes,
are required to be productive \cite{Coquand93} in the sense that every
finite unfolding can be computed in a finite number of steps.
In Agda's encoding of coinductive types using ``musical notation''
\NV{Citation? What about Coq?}, productivity is achieved using a strict method: in the
definition of corecursive functions, the recursive
call is required to appear directly under
the application of a constructor. This restriction
typically makes programming with processes cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
\NV{Cite Huges?} is a type annotated with the number of unfolding that
elements of this type can undergo. Sized types have been implemented
in Agda \cite{A-sized,AVW-normalization} and can be used in
combination with coinductive records \NV{Cite Copatterns?} for the
specification of coinductive types.  Guarded recursion \NV{Cite
  Nakano?} is a different approach where the type system is enhanced
with a modality, called ``later'' and written \IC{▻}, encoding time
delay in types. The later modality comes with a general fixpoint
combinator for programming with productive recursive functions and
allows the specification of guarded recursive types, \eg the
type of guarded streams satisfying the isomorphism \F{Str}
\Ar{A} \F{≃} \Ar{A} \F{×} \IC{▻} (\F{Str} \F{A}). The tail
of a guarded stream is available only one time step ahead from now, as
evidenced by the application of the later modality to \F{Str}
\F{A}. Coinductive types can be constructed from guarded recursive
types in type theories that additionally support clock variables and
universal quantification over them
\cite{atkey2013productive,BahrGM17}. An implementation of
calculi with guarded recursion is currently missing.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

>>>>

Dependent type theory is a very expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda and Lean.  It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Elements of these types, which we call processes,
are required to be productive \cite{Coquand93} in the sense that every
finite unfolding can be computed in a finite number of steps.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10} \NV{What about Coq?}, productivity is achieved using a strict method: in the
definition of corecursive functions, the recursive
call is required to appear directly under
the application of a constructor. This restriction
typically makes programming with processes cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
\cite{Hughes96} is a type annotated with the number of unfolding that
elements of this type can undergo. Sized types have been implemented
in Agda \cite{A-sized,AVW-normalization} and can be used in
combination with coinductive records for the specification of
coinductive types \cite{Copatterns}.  Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \F{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \F{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus with guarded recursion.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Sized types and guarded recursion are different solution to the same
problem, but a thorough study of the relation between these two
disciplines is still missing. In this paper we take a first step in
this direction by showing how guarded recursion can be simulated in
Agda using sized types.
